Safe Transferable Regions

### 2016-07-02 paper/type-inference.tex
soundness, completeness, examples

The type inference algorithm is sound:

@@

Furthermore, the second phase of the type inference algorithm is complete as well.
If there is any possible instantiation (of the region variables and predicate variables) of
the elaborated program produced by the first phase that type checks, then the second phase
will identify it.

Incompleteness exists only in a few choices made during the first phase (elaboration phase), as
explained below.

@@

\textbf{TODO: Update example discussion}

\subsection{Example}

\paragraph{Constraints Example 1} Consider the $\C{Pair}$ class
template from \S\ref{sec:fb-templatization}. Following constraints are
generated during its elaboration (Constrains are identified with
$\mathbf{c_i}$'s. Some trivial constraints, such as $\rho_4 \in
\rhoenv_0$ and $\rho_5 \in \rhoenv_1$, where $\rhoenv_0 =
\{\rhoalloc_0,\rho_{0-4}\}$ and $\rhoenv_1 = \rhoenv_0 \cup
\{\rho_5\}$, have been elided): 
\begin{smathpar}
\begin{array}{l}
  \csid{1} \tywf{\rhoenv_0}{\varphi_0} \qquad
  \csid{2} \isvalid{\varphi_0}{\rho_0 \outlives \rhoalloc_0 \conj \rho_1
     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0} \\
  \csid{3} \isvalid{\varphi_0}{\rho_2 = \rho_0} \spc
  \csid{4} \isvalid{\varphi_0}{\rho_3 = \rho_1} \\
  \csid{5} \isvalid{\varphi_0 \conj \varphi_1} {\rho_5 = \rho_0}\spc
  \csid{6} \tywf{\rhoenv_1}{\varphi_1} \qquad
\end{array}
\end{smathpar}

\paragraph{Constraints Example 2} Let us add to the \C{Pair} class a
contrived method $\C{alt}$ that accepts a \C{Region} object \C{r}, a
\C{Pair} object \C{q}, and an object \C{y}. It assigns
\C{y} to \C{fst} and \C{snd} fields of \C{q}, and calls itself
recursively with the same region, a new \C{Pair} object allocated in a
local region, and an object referred by the \C{snd} field of the
pair inside the region. \C{alt} never terminates.  Elaboration phase
elaborates the method to the following region-annotated
definition\footnote{In reality, elaboration uses new region variables
as parameters to the method calls, and then generates
constraints that unify them with actuals. In our examples, to avoid
clutter due to trivial constraints,we coalesced both steps and show
the actuals instead.}(The original definition of \C{alt} can be
obtained by erasing all the region annotations from the elaborated
version):
% \begin{codejava}
% unit alt<$\rhoalloc_2$,$\rho_{6-10}$ | $\varphi_2$>(Pair<A,A><$\rho_6$,$\rho_7$,$\rho_8$> p,
%                           A<$\rho_9$> x, A<$\rho_{10}$> y) {
%   p.fst = x; p.snd = y; 
%   this.alt<$\rhoalloc_2$,$\rho_7$,$\rho_8$,$\rho_{10}$,$\rho_9$>(A,y,x);
% }
% \end{codejava}
\begin{codejava}
unit alt<$\rhoalloc_2$,$\rho_{6-9}\,$|$\,\varphi_2$>(Region<Pair><$\toprgn$> r, 
                Pair<$\rho_{6-8}$> q, Object<$\rho_{9}$> y) {
  q.fst := y; q.snd := y; 
  open r as p@$\rgn_0$ in
    letregion $\rgn_1$ in
      let x = new Pair<$\rgn_1$,$\rgn_0$,$\rgn_0$>
                      (p.fst,p.fst) in
        alt<$\rgn_{1}$,$\rgn_{1}$,$\rgn_{0}$,$\rgn_{0}$,$\rgn_{0}$>(r,x,p.snd)
}
\end{codejava}
% Note that, to avoid clutter, we have already resolved appropriate
% region arguments to the recursive call, instead of introducing new
% region variables and generating equality constraints on them. 
Constraints generated during the elaboration are shown below
(let $\rhoenv_2 = \{\rhoalloc_0,\rho_{0-4},\rhoalloc_2,\rho_{6-9}\}$ ):
\begin{smathpar}
\begin{array}{l}
\csid{7} \tywf{\rhoenv_2}{\varphi_2}\qquad
\csid{8} \isvalid{\varphi_1 \conj \varphi_2}
    {\rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6} \\
\csid{9} \isvalid{\varphi_1 \conj \varphi_2}{\rho_7 = \rho_9} \qquad
\csid{10} \isvalid{\varphi_1 \conj \varphi_2}{\rho_{8} = \rho_9} \\
\csid{11} \isvalid{\varphi_1 \conj \varphi_2 \conj \rgn_0 \outlives
\rgn_1} 
    {[\rgn_1/\rhoalloc_2][\rgn_1/\rho_6][\rgn_0/\rho_{7-9}]\varphi_2}
\end{array}
\end{smathpar}

% \paragraph{Constraints Example 1} The original source of all
% outlives constraints is the validity constraint $C_3$ in $\elabClass$
% function (Fig.~\ref{fig:fb-elabmeth}). In context of the $\C{Pair}$
% class template from \S\ref{sec:fb-templatization}, the constraint is
% as following:
% \begin{center}
%   \(\isvalid{\varphi_0}{\rho_0 \succeq \rhoalloc_0 \conj \rho_0
%   \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0}\)
% \end{center}

>>>>

We now state the soundness theorem:

@@

We do not present a formal proof of the theorem. The proof essentially requires
establishing the following lemma inductively: if
$
 \exprok {\exptycx{\ralloc}{\env}}{e}{\tau}{C}
$
and $(\regionSubstFn,\predSubstFn)$ satisfies $C$, then
$
 \hastyp {\exptycx{\ralloc}{\env}}{e[\regionSubstFn,\predSubstFn]}{\tau}
$,
where $e[\regionSubstFn,\predSubstFn]$ indicates $e$ with the substitutions
in the solution $(\regionSubstFn,\predSubstFn)$,
as well as analogous lemmas for type well-formedness.

However, in this context, completeness is more interesting than soundness.
(Note that an unsound inference algorithm can be made sound by simply type-checking
the resulting program.)
We believe that the converse of the above lemma can also be established:
any substitution for the region variables and the predicate variables that makes the
expression well-typed will also satisfy the generated constraints.
Thus, if there is any possible instantiation (of the region variables and predicate variables) of
the elaborated program produced by the first phase that type checks, then the generated
constraints are satisfiable, and the second phase will identify it.

However, the entire type inference algorithm is not complete, because of some heuristic
choices made during the first phase (elaboration phase), as explained below.

@@

%\textbf{TODO: Update example discussion}
%
%\subsection{Example}
%
%\paragraph{Constraints Example 1} Consider the $\C{Pair}$ class
%template from \S\ref{sec:fb-templatization}. Following constraints are
%generated during its elaboration (Constrains are identified with
%$\mathbf{c_i}$'s. Some trivial constraints, such as $\rho_4 \in
%\rhoenv_0$ and $\rho_5 \in \rhoenv_1$, where $\rhoenv_0 =
%\{\rhoalloc_0,\rho_{0-4}\}$ and $\rhoenv_1 = \rhoenv_0 \cup
%\{\rho_5\}$, have been elided): 
%\begin{smathpar}
%\begin{array}{l}
%  \csid{1} \tywf{\rhoenv_0}{\varphi_0} \qquad
%  \csid{2} \isvalid{\varphi_0}{\rho_0 \outlives \rhoalloc_0 \conj \rho_1
%     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0} \\
%  \csid{3} \isvalid{\varphi_0}{\rho_2 = \rho_0} \spc
%  \csid{4} \isvalid{\varphi_0}{\rho_3 = \rho_1} \\
%  \csid{5} \isvalid{\varphi_0 \conj \varphi_1} {\rho_5 = \rho_0}\spc
%  \csid{6} \tywf{\rhoenv_1}{\varphi_1} \qquad
%\end{array}
%\end{smathpar}
%
%\paragraph{Constraints Example 2} Let us add to the \C{Pair} class a
%contrived method $\C{alt}$ that accepts a \C{Region} object \C{r}, a
%\C{Pair} object \C{q}, and an object \C{y}. It assigns
%\C{y} to \C{fst} and \C{snd} fields of \C{q}, and calls itself
%recursively with the same region, a new \C{Pair} object allocated in a
%local region, and an object referred by the \C{snd} field of the
%pair inside the region. \C{alt} never terminates.  Elaboration phase
%elaborates the method to the following region-annotated
%definition\footnote{In reality, elaboration uses new region variables
%as parameters to the method calls, and then generates
%constraints that unify them with actuals. In our examples, to avoid
%clutter due to trivial constraints,we coalesced both steps and show
%the actuals instead.}(The original definition of \C{alt} can be
%obtained by erasing all the region annotations from the elaborated
%version):
%% \begin{codejava}
%% unit alt<$\rhoalloc_2$,$\rho_{6-10}$ | $\varphi_2$>(Pair<A,A><$\rho_6$,$\rho_7$,$\rho_8$> p,
%%                           A<$\rho_9$> x, A<$\rho_{10}$> y) {
%%   p.fst = x; p.snd = y; 
%%   this.alt<$\rhoalloc_2$,$\rho_7$,$\rho_8$,$\rho_{10}$,$\rho_9$>(A,y,x);
%% }
%% \end{codejava}
%\begin{codejava}
%unit alt<$\rhoalloc_2$,$\rho_{6-9}\,$|$\,\varphi_2$>(Region<Pair><$\toprgn$> r, 
%                Pair<$\rho_{6-8}$> q, Object<$\rho_{9}$> y) {
%  q.fst := y; q.snd := y; 
%  open r as p@$\rgn_0$ in
%    letregion $\rgn_1$ in
%      let x = new Pair<$\rgn_1$,$\rgn_0$,$\rgn_0$>
%                      (p.fst,p.fst) in
%        alt<$\rgn_{1}$,$\rgn_{1}$,$\rgn_{0}$,$\rgn_{0}$,$\rgn_{0}$>(r,x,p.snd)
%}
%\end{codejava}
%% Note that, to avoid clutter, we have already resolved appropriate
%% region arguments to the recursive call, instead of introducing new
%% region variables and generating equality constraints on them. 
%Constraints generated during the elaboration are shown below
%(let $\rhoenv_2 = \{\rhoalloc_0,\rho_{0-4},\rhoalloc_2,\rho_{6-9}\}$ ):
%\begin{smathpar}
%\begin{array}{l}
%\csid{7} \tywf{\rhoenv_2}{\varphi_2}\qquad
%\csid{8} \isvalid{\varphi_1 \conj \varphi_2}
%    {\rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6} \\
%\csid{9} \isvalid{\varphi_1 \conj \varphi_2}{\rho_7 = \rho_9} \qquad
%\csid{10} \isvalid{\varphi_1 \conj \varphi_2}{\rho_{8} = \rho_9} \\
%\csid{11} \isvalid{\varphi_1 \conj \varphi_2 \conj \rgn_0 \outlives
%\rgn_1} 
%    {[\rgn_1/\rhoalloc_2][\rgn_1/\rho_6][\rgn_0/\rho_{7-9}]\varphi_2}
%\end{array}
%\end{smathpar}
%
% \paragraph{Constraints Example 1} The original source of all
% outlives constraints is the validity constraint $C_3$ in $\elabClass$
% function (Fig.~\ref{fig:fb-elabmeth}). In context of the $\C{Pair}$
% class template from \S\ref{sec:fb-templatization}, the constraint is
% as following:
% \begin{center}
%   \(\isvalid{\varphi_0}{\rho_0 \succeq \rhoalloc_0 \conj \rho_0
%   \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0}\)
% \end{center}
